---
title: "tidyr tutorial"
author: "Joseph V. Casillas"
date: 'Updated: `r Sys.time()`'
output:
  html_document:
    highlight: textmate
    includes:
      after_body: includes/after_body.html
      before_body: includes/before_body.html
      in_header: includes/in_header.html
    theme: spacelab
---

```{r echo=FALSE}
library(knitr)
read_chunk('tidyr.R')
```

```{r, 'load', echo=FALSE, message=FALSE}
```


## Overview

In order to facilitate the data analysis pipeline, it is crucial 
to have `tidy data`. What this means is that **every column in 
your data frame represents a variable and every row represents an observation**. This is also referred to as *long format* (as 
opposed to wide format).

`tidyr` is a package that provides useful functions for converting 
raw data into tidy data. This is typically the first step in the 
data analysis pipeline after you have collected your data.

```{r, 'pipeline1', echo=FALSE, fig.align='center', eval=TRUE}
```

This tutorial will focus on step 2 of the process. The main verbs 
we will use are:

- `gather()` and `spread()` in order to convert between long and 
wide data
- `separate()` can split up a single column into servaral variables 
and is more commonly used in conjunction with `gather()` for 
linguistic research (i.e. when separating columns in praat).


## gather()

- used to make wide data long
- takes columns, and **gathers** them into key-value pairs
- `gather(df, newVar1, newVar2, vector1, vector2)`

```{r}
library(tidyr); library(dplyr)

set.seed(1)
tidyr.ex <- data.frame(
  participant = c("p1", "p2", "p3", "p4", "p5", "p6"), 
  info = c("g1m", "g1m", "g1f", "g2m", "g2m", "g2m"),
  day1score = rnorm(n = 6, mean = 80, sd = 15), 
  day2score = rnorm(n = 6, mean = 88, sd = 8)
)

print(tidyr.ex)

tidyr.ex %>%
  gather(day, score, c(day1score, day2score))
```

Essentially we took the columns `day1score` and `day2score`, which represent the variable `day` and the variable `score`, and *gathered* them. Why? Remember that tidy data has one column for each variable and one row for each observation. The numbers in the two columns we changed were observations, thus they should each get their own row. 

## spread()

- This is a compliment of `gather()`. The `spread() verb takes different levels of a factor and spreads them out into different columns. This means we can convert from long data to wide.
- `spread(df, var1, var2)


```{r, eval=TRUE}
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  spread(day, score)
```

Now we are back to how we started. 

## separate()

- Takes values inside a column and separates them.
- Ex. mg1old > m g1 old
- `separate(df, col, into, sep)`

Consider the column `info` of our fake data. You can probably guess 
what observations represent. How many variables are there? Take a 
second to think about it if it doesn't jump out at you. The answer is 
2. `g1` and `g2` appear to be a grouping variable (g = group) and `m` 
`f` is an indication of gender. Because there are two separate 
variables, there should be two columns in the data frame... one for 
`group` and one for `gender`. 


```{r, eval=TRUE}
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  separate(col = info, into = c("group", "gender"), sep = 2)
```

## unite()

- Unite does the opposite of spread. In my experience, this is not 
something that needs to be done very often. 
- unite(df, newVarName, col1, col2)

```{r, eval=TRUE}
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  separate(col = info, into = c("group", "gender"), sep = 2) %>%
  unite(infoAgain, group, gender)
```

Now that our data are tidy (using just the ```gather()``` and ```separate()``` verbs), we can plot and analyze it. 

```{r, fig.align='center'}
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  separate(col = info, into = c("group", "gender"), sep = 2) %>%
  ggplot(aes(x = day, y = score)) + 
  geom_point() + 
  facet_wrap(~ group) +
  geom_smooth(method = "lm", aes(group = 1), se = F)
```


These are the essential verbs used for tidying data. There are other 
commands that can be useful, but mainly they are different takes on 
the ones we have covered here (i.e. `extract()` and `unite()`, which 
are similar to `separate()` and `gather()`, respectively, but use 
regex). 






